---
title: "Spatial Hotspots and Seasonal Shifts in Juvenile Loggerhead Sea Turtle Telemetry Data (2002–2005)"
https://github.com/uilliam-c/Connolly_Liam_ENE872_FinalProject
author: "Liam Connolly"
output: bookdown::pdf_document2
  toc: false
    number_selections: true
---
\newpage
\tableofcontents 
\newpage
\listoftables 
\newpage
\listoffigures 
\newpage

```{r setup, include=FALSE}
# Set your working directory
knitr::opts_chunk$set(echo = FALSE, message = FALSE, warning = FALSE)
# Load your packages
library(tidyverse)
library(lubridate)
# Set your ggplot theme
theme_set(theme_bw())
# Load your datasets
turtle_points <- readr::read_csv("../data_raw/turtlepoints.csv",
  show_col_types = FALSE)

modifiedturtle_points <- turtle_points %>%
  mutate(date_time =ymd_hms(date_time, tz = "UTC"),
  month = month(date_time), 
  season = ifelse(month %in% c(12, 1, 2), "Winter",
           ifelse(month %in% c(3, 4, 5), "Spring",
           ifelse(month %in% c(6, 7, 8), "Summer", "Fall"))),
  season = factor(season, levels = c("Winter", "Spring", "Summer", "Fall")),
  classitem = factor(classitem)) %>%
  filter(!is.na(latitude), !is.na(longitude), !is.na(date_time))
modifiedturtle_points <- modifiedturtle_points %>%
  mutate(lon_wrap = ifelse(longitude > 180, longitude - 360, longitude))
if (!file.exists("../data_processed/turtle_points_clean.csv")) {
  write_csv(modifiedturtle_points, "../data_processed/turtle_points_clean.csv")
}
```


# 1. Rationale
According to the online database, from which I collected my data, juvenile 
loggerhead turtles (Caretta caretta) move around frequently. Due to their 
juvenile phase of life, they are highly mobile yet, what isn't known is how or 
why they move to certain places in different seasons or ocean regions. 
From a spatial perspective, we don't have answers. Being able to identify where 
and when these turtles move/are found can help better understand their habitats.
It would be awesome to discover "hotspots" of more concentrated turtle presence 
as that may guide additional ecological analyses in the future. With the data 
available, it should be possible to assemble a general sense of the spatial 
patterns these loggerheads exhibit.

The background of this dataset relies on using satellite tracking locations
where each record present is indicative of a geo-referenced observation of a 
juvenile loggerhead. Each record comes with latitude/longitude, an individual 
identifier (series_id), and a timestamp (date_time). It is likely the dataset
will include repeated observations from the same loggerhead over the course of
the study (2002-2005), the data is still relevant for basic analyses. I will be 
focusing on examing seasonal distribution and spatial concentration for the 
loggerheads. Finally, the data contains a location "class" code (classitem) and 
some precision fields to help evaluate the patterns of data quality. 

# 2. Research Questions
Research Topic 1) Seasonal distribution Shifts
Research Question: Do juvenile loggerhead sea turtle locations (latitude)
differ among seasons (Winter, Spring, Summer, Fall)?
Hypothesis: Mean latitude differs among seasons.
Rationale for hypothesis: By nature, seasonal changes in ocean conditions such 
as temperature, current direction, etc should influence where juvenile turtles 
are found. If this were true that juvenile loggerheads go to different places
throughout the year, or if observation density varies by the different seasons
in varying areas, then there should be differences in latitude 
among seasons.

Research Topic 2) Spatial Hotspots of Presence
Research Question : Are loggerhead turtles locations concentrated in a few 
specific geographic areas (“Hotspots”) or are they more evenly distributed?
Hypothesis: The locations are spatially clustered, with a small amount of 
places accounting for a disproportionate amount of loggerhead observations.
Rationale: It isn't uncommon for animals, when they find a favorable place to 
live with good food availability or water temperature, to appear repeatedly in 
the same area. Therefore, if I can discover repeated observations of 
loggerheads in the same types of areas, it should allow for the creation of 
measurable hotspots in a spatial setting.

Research Topic 3) Classitem Differences by Location Class
Research Question: Does the distribution of classitem categories
differ by season?
Hypothesis: The variety of observations in each classitem category
will differ among seasons (Winter, Spring, Summer, Fall),
indicating that the mix of class item codes will also vary over time.
Rationale: While examining the dataset, I came across the columns with a largely
unexplained categorical code (I think its categorical) that wasn't mentioned
anywhere in the public database. That code being classitem. It appears the
classitem has some variablity between observations and that the recorded 
observations have no even distribution across the study. In theory, the
data collected via satellite would be under different conditions including 
seasonal shifts in loggerhead behavior so, it is possible that the proportion
of observations in each class item category differs among seasons. 
\newpage

# 3. Dataset Information
This dataset originates from OBIS-SEAMAP. Specifically, the dataset is called
“SWFSC juvenile loggerhead sea turtle tracking 2002–2005” 
(Dataset ID 758; Version 1.0.1; internal DOI: 10.82144/4774e438). Actual URL is:
https://seamap.env.duke.edu/dataset/758
The metadata explains in further detail that the source is satellite tracking 
data of the juvenile loggerhead turtle (Caretta caretta). The dataset has data 
from October 14, 2002 to July 7, 2005. Location coordinates are provided in 
decimal degrees (coordinate system: WGS84), and the spatial extent 
for the dataset is approximately 20.95–41.913° N latitude and 
153.571–246.713° E longitude. 

Each row in the points file is a single location observation for an 
individually separate turtle. Also included with an observation is the latitude
and longitude (represented in decimal degrees) and observation date/time. There
are some limitations as publically downloading the data limits me to core fields
including date/time, latitude/longitude, and species information. It is unclear
if the categories I can't download will influence the data analyses I conduct. 
Reported precision fields (lprecision and tprecision) were constant in this
dataset, so precision-based comparisons were not informative. 
\newpage

# 4. Data Wrangling
I imported the points CSV into R-studio Online and converted the 
date_time column to date_time in UTC. I removed any records with missing
latitude, longitude, or timestamp information. In this case, just one record was
removed. Not knowing this in advance that only one record would be removed,
I intended to ensure all analyses had a complete set of spatial observations
recorded to remove outlier data. Additionally, I created a categorical season 
variable from the months included in date_time. Winter was defined as Dec–Feb,
Spring = Mar–May, Summer = Jun–Aug, and Fall = Sep–Nov. Finally, I converted
class_item to a categorical variable for later group comparison.

```{r}
table1 <- tibble::tibble(
  Detail = c("Data source",
             "Retrieved from",
             "Dataset title",
             "Dataset ID",
             "DOI",
             "Date range",
             "Coordinate system"),
  Description = c("OBIS-SEAMAP",
                  "https://seamap.env.duke.edu/dataset/758",
                  "SWFSC juvenile loggerhead sea turtle tracking 2002–2005",
                  "758",
                  "https://doi.org/10.82144/4774e438",
                  "2002-10-14 to 2005-07-07",
                  "WGS84; decimal degrees")
)

knitr::kable(table1, caption = "Dataset information.")

table2_vertical <- tibble::tibble(
  Detail = c("Number of observations", "Number of turtles", "Start date-time", 
          "End date-time", "Minimum latitude", "Maximum latitude", 
          "Minimum longitude", "Maximum longitude", "Median lprecision"),
  Value = c(nrow(modifiedturtle_points),
            dplyr::n_distinct(modifiedturtle_points$series_id),
            as.character(min(modifiedturtle_points$date_time, na.rm = TRUE)),
            as.character(max(modifiedturtle_points$date_time, na.rm = TRUE)),
            min(modifiedturtle_points$latitude, na.rm = TRUE),
            max(modifiedturtle_points$latitude, na.rm = TRUE),
            min(modifiedturtle_points$longitude, na.rm = TRUE),
            max(modifiedturtle_points$longitude, na.rm = TRUE),
            median(modifiedturtle_points$lprecision, na.rm = TRUE))
)

knitr::kable(table2_vertical, caption = "Data structure and 
Basic ranges for Key Variables Used.")

```
\newpage

# 5. Exploratory Analysis 
This analysis of the data used gives a general summary of the spatial
structure of the juvenile loggerhead turtle location dataset prior to testing
my three hypotheses. First, I use maps of latitude and longitude to describe
the geographic extent of observations and then I try to do a summary
over time to get a sense of the patterns of observation across months and
seasons. To inquire the spatial concentration for question 2, I tried to express
a visualization of point density to identify whether sightings appear
clustered into potential “hotspots.” Additionally, I was curious to see if I
could separate the observations between individual turtles (via series_id) to
find repeated sightings. Finally, I try to express latitude by season
(Winter, Spring, Summer, Fall) to address the seasonal questions from the first
question and to compare the reported location precision (lprecision) with
classitem categories. This is to determine if the differences in positional 
precision could influence the spatial clustering that I was curious of in my
third research question.
```{r}
missingdata_table <- modifiedturtle_points %>%
  summarise(
    n_total = n(),
    missing_lat = sum(is.na(latitude)),
    missing_lon = sum(is.na(longitude)),
    missing_date_time = sum(is.na(date_time)),
    missing_lprecision = sum(is.na(lprecision)),
    missing_classitem = sum(is.na(classitem))
  ) %>%
  mutate(across(everything(), as.character)) %>%
  tidyr::pivot_longer(everything(), names_to = "Detail", values_to = "Value")
knitr::kable(missingdata_table, caption = "Representing Missing Data 
             with Key Variables.")
```
This table shows, after filtering, that there are no missing values within the 
key variables used for analysis (latitude, longitude, timestamp, precision, 
and class code). With this in mind, further plots and tests are being conducted
off a complete dataset of 6861 location observations.
```{r map, fig.cap="Spatial Patterns for Turtles.", fig.width=7, fig.height=4, out.width="0.9\\linewidth", fig.align="center"}
ggplot(modifiedturtle_points, aes(x = lon_wrap, y = latitude)) +
  geom_point(alpha = 0.3) +
  labs(x = "Longitude", y = "Latitude")
```
Figure shows the geographic extent of juvenile loggerhead turtle
locations within this dataset. 

```{r time, fig.cap="Number of location observations over time (by month)."}
time_counts <- modifiedturtle_points %>%
  mutate(month_start = floor_date(date_time, unit = "month")) %>%
  count(month_start)
ggplot(time_counts, aes(x = month_start, y = n)) +
  geom_col() +
  scale_x_date(date_breaks = "6 months", date_labels = "%Y-%m") +
  labs(x = "Month", y = "Number of observations")
```
This Figure shows sampling intensity across the duration of the study period.

```{r time, fig.cap="Number of location observations over time (by month)."}
time_counts <- modifiedturtle_points %>%
  mutate(month_start = floor_date(date_time, unit = "month")) %>%
  count(month_start)

ggplot(time_counts, aes(x = month_start, y = n)) +
  geom_col() +
  scale_x_date(date_breaks = "6 months", date_labels = "%Y-%m") +
  labs(x = "Month", y = "Number of observations")
```
This figure shows that separate turtles provide different numbers
of observations.

```{r lat_season, fig.cap="Latitude by season."}
ggplot(modifiedturtle_points, aes(x = season, y = latitude)) +
  geom_boxplot() +
  labs(x = "Season", y = "Latitude")
```

This figure shows latitude spread across Winter, Spring, Summer, and Fall.

```{r hot, fig.cap="Exploratory point density (1-degree grid)."}
hotgrid <- modifiedturtle_points %>%
  mutate(
    lat_bin = floor(latitude),
    lon_bin = floor(lon_wrap)
  ) %>%
  count(lat_bin, lon_bin)

ggplot(hotgrid, aes(x = lon_bin, y = lat_bin, fill = n)) +
  geom_tile() +
  labs(x = "Longitude (binned)", y = "Latitude (binned)", fill = "Count")
```
This figure inquires and suggests areas of concentrated observations
that motivate the hotspot analysis for Q2.

```{r class_season, fig.cap="Proportion of observations in the most common classitem categories by season."}
top_class <- modifiedturtle_points %>%
  count(classitem, sort = TRUE) %>%
  slice_head(n = 6) %>%
  pull(classitem)

modifiedturtle_points %>%
  mutate(classitem2 = ifelse(classitem %in% top_class, as.character(classitem), "Other")) %>%
  count(season, classitem2) %>%
  ggplot(aes(x = season, y = n, fill = classitem2)) +
  geom_col(position = "fill") +
  labs(x = "Season", y = "Proportion of observations", fill = "classitem")
```
This figure shows the varying proportionate composition of classitem 
categories across different seasons. While many classitem categories are
shown, a few make up most observations. The relative proportions appear similar
between seasons.
\newpage

# 6. Analysis


## Question 1: Do juvenile loggerhead sea turtle locations (latitude) differ 
## among seasons (Winter, Spring, Summer, Fall)?
```{r rq1_prep, include=FALSE}
lat_season_means <- modifiedturtle_points %>%
  group_by(series_id, season) %>%
  summarise(
    mean_lat = mean(latitude, na.rm = TRUE),
    n_obs = n(),
    .groups = "drop"
  )
```

```{r rq1_fig, fig.cap="Mean latitude by season (summarized per turtle)."}
ggplot(lat_season_means, aes(x = season, y = mean_lat)) +
  geom_boxplot() +
  labs(x = "Season", y = "Mean latitude (per turtle)")
```
```{r rq1_test, include=FALSE}
rq1_aov <- aov(mean_lat ~ season, data = lat_season_means)
rq1_aov_summary <- summary(rq1_aov)
rq1_season_summary <- lat_season_means %>%
  group_by(season) %>%
  summarise(
    n_turtles = n(),
    mean_lat = mean(mean_lat, na.rm = TRUE),
    sd_lat = sd(mean_lat, na.rm = TRUE),
    .groups = "drop"
  )
rq1_season_summary
```
After being able to summarize latitude by individual turtle occurrences and
season, mean latitude differed significantly among seasons (ANOVA; df = 3,
58; F = 38.3; p = 8.97 × 10⁻¹⁴ or p<0.0001). This figure clearly shows
seasonal separation in per-turtle mean latitude. With that said, these results
support my hypothesis that juvenile loggerhead latitude varies by season
in this dataset. Mean latitude was highest in Fall (36.89°N) and lowest
in Spring (32.02°N). This is consistent with a seasonal shift in latitude.

## Question 2: Are loggerhead turtles locations concentrated in a few 
## specific geographic areas (“Hotspots”) or are they more evenly distributed?

```{r rq2-hotspots, echo=FALSE, message=FALSE, warning=FALSE, fig.cap="Top 20 most-used 1-degree grid cells (counts of turtle locations per grid cell)."}
grid_counts <- modifiedturtle_points %>%
  mutate(
    lat_cell = floor(latitude),
    lon_cell = floor(lon_wrap)
  ) %>%
  count(lat_cell, lon_cell, name = "n_cell") %>%
  arrange(desc(n_cell))

n_total <- nrow(modifiedturtle_points)
top10_share <- sum(grid_counts$n_cell[1:min(10, nrow(grid_counts))]) / n_total

top_cells <- grid_counts %>%
  slice_head(n = 20) %>%
  mutate(cell = paste0("(", lat_cell, ", ", lon_cell, ")")) %>%
  arrange(n_cell) %>%
  mutate(cell = factor(cell, levels = cell))

ggplot(top_cells, aes(x = cell, y = n_cell)) +
  geom_col() +
  coord_flip() +
  labs(x = "Grid cell (lat, lon)", y = "Number of locations")

top10_table <- grid_counts %>% slice_head(n = 10)
knitr::kable(top10_table, caption = "Top 10 Most-Used 1-Degree Grid cells.")
```
To evaluate RQ2, all loggerhead locations were summarized into 1-degree
latitude–longitude grid cells and the number of observations per cell
was counted. Under an even spatial distribution, counts would be similar
across grid cells. However, hotspot clustering would be supported if a small
subset of cells had a disproportionate share of the locations. In this case, the
results show clear spatial concentration. The top 10 grid cells contained `r scales::percent(top10_share, accuracy = 0.1)` of all locations 
(Figure \@ref(fig:rq2-hotspots)). The highest-use cells had significantly higher
counts of locations than the others. These findings support the RQ2 hypothesis I
made that loggerhead turtle locations are clustered into a small number of
hotspots rather than evenly distributed across the study area.

Given the possible data variability, as a data-quality check, longitudes were
standardized to −180 to 180 where (min = −180.0, max = 180.0). Of those,
15.4% of locations observed occurred at positive longitudes. 

## Question 3: Does the distribution of classitem categories differ by season?
```{r rq3-classitem-season, echo=FALSE, message=FALSE, warning=FALSE, fig.cap="Classitem composition by season (top categories shown; remaining grouped as 'Other')."}
N_top <- 8

top_levels <- modifiedturtle_points %>%
  count(classitem, sort = TRUE) %>%
  slice_head(n = N_top) %>%
  pull(classitem)

rq3_df <- modifiedturtle_points %>%
  mutate(classitem2 = ifelse(classitem %in% top_levels, as.character(classitem), "Other"))

rq3_prop <- rq3_df %>%
  count(season, classitem2) %>%
  group_by(season) %>%
  mutate(prop = n / sum(n)) %>%
  ungroup()

ggplot(rq3_prop, aes(x = season, y = prop, fill = classitem2)) +
  geom_col() +
  labs(x = "Season", y = "Proportion", fill = "classitem")

rq3_table <- rq3_prop %>%
  select(season, classitem2, prop) %>%
  tidyr::pivot_wider(names_from = season, values_from = prop)

knitr::kable(
  rq3_table, digits = 3,
  caption = "Proportion of observations in each classitem category by season (top categories + Other)."
) 
```

```{r rq3-biggest-changes, echo=FALSE, message=FALSE, warning=FALSE, fig.cap="Largest seasonal changes in classitem composition (max minus min seasonal proportion)."}
rq3_changes <- rq3_prop %>%
  group_by(classitem2) %>%
  summarise(
    min_prop = min(prop),
    max_prop = max(prop),
    range_prop = max_prop - min_prop,
    .groups = "drop"
  ) %>%
  arrange(desc(range_prop))

rq3_changes_top10 <- rq3_changes %>% slice_head(n = 10)

knitr::kable(
  rq3_changes_top10, digits = 3,
  caption = "Top 10 classitem categories with the largest seasonal change in proportion (top categories + Other)."
)
```
For RQ3 (research question 3), I compared the composition of classitem
categories across seasons. Due to classitem containing many categories, I
only displayed the most frequent categories and all remaining categories were
noted as “Other”. The proportion plot shows that classitem composition varied
across seasons. For example, the “Other” category decreases from 0.404 in Winter
to 0.206 in Fall. By contrast, while classitem 5 increases from 0.059 in Winter
to 0.168 in Summer, Classitem 21 also increases toward Fall 
(0.080 in Winter to 0.141 in Fall). The tables also summarize the seasonal
proportions for the categories and identify the categories with the
largest seasonal changes (those being classitem 5 and 21). This analysis is
descriptive and does not include a test of statistical significance.

In review, these results match my RQ3 hypothesis that classitem composition
would differ by season. This is because multiple categories,
as well as the non listed “Other” group, change noticeably across Winter, 
Spring, Summer, and Fall. However, it is important to note as I did above, that
because this analysis is descriptive, the results support the hypothesis based
on observed patterns. As I did not run a test of statistical significance, its
possible for the results to be different. 
\newpage

# Summary and Conclusions


\newpage

# References
<add references here if relevant, otherwise delete this section> 
